---
title: "Module 5"
author: "Data Pirates"
date: "2022-10-05"
output: html_document
bibliography: DataPirates.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
```

# Introduction
For this project, our gruop will be investigating whether a person's body temperature varies with different levels of activity. According to a study by @lim2008human, physical activity causes a substantial increase in metabolic heat production, thus increasing body temperature. To further investigate this, the temperature of 4 different subjects during their rest time and after (fill in time once we record data in case we don't do 5 mins) minutes of activity will be recorded. 

# Methods
In order to determine the relationship between body temperature and activity level, each individual had their body temperature (°C) measured from their armpit for (however long we end up doing). After that has been recorded, the individual will proceed to do jumping jacks for (however long) and their body temperature from their armpit will be recorded a second time. This process will be done to all four individuals and their data will be recorded accordingly. 

# Results
```{r,"adding in our own data"}
library(tidyverse)
dat.f <- list.files(pattern="couch|active.csv")
dat.l <- list()
```
```{r, "for loop"}
for (i in dat.f){
  met.dat <- unlist(strsplit(i,"_"))
  who <- met.dat[1]
  activity <- gsub(".csv","",met.dat[2])
  dat.l[[i]] <- read_csv(i,col_names = TRUE,col_types = NULL)%>%
    mutate("who" = who, "activity" = activity)
}
```
```{r, "making data tables"}
dat <- do.call(rbind,dat.l)

```
```{r,"[plotting tables",fig.cap= "Individuals' varying temperatures over time during different states of activity; one in a state of rest (couch) and the other, during physical activity (active)"}
plot <- dat%>%
  ggplot(aes(x=activity,y=degC))+geom_boxplot()+stat_summary(fun=mean,geom="point",size=3)+geom_errorbar(aes(ymin=degC-sd(degC), ymax=degC+sd(degC)),width=.002)+ylab("Body Temperature (Degrees Celsius)")+facet_grid(.~who)
print(plot)
```

Based off of the findings in the plot above, it appears that body temperature fluctuation during physical activity compared with temperature during rest does not drastically vary, for a majority of individuals that were tested. This suggests that physical activity does not have a direct affect on changing body temperature. 

# Discussion

@fortney1985exercise
@gleeson1998temperature

# Author Contribution
Chris - Wrote Introduction and Methods paragraphs
Alexa - Recorded, Saved, and Properly Named Data
Ellie - Put metadata in table and created results graph; wrote results narrative on findings. 
Tori - 
# References
